---
title: "Differential Gene Expression in Human Gingival Tissue During Periodontitis"
author: "Andrea Torres"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
Periodontitis is a chronic inflammatory disease that affects the supporting tissues of the teeth. This project analyzes RNA-seq data from human gingival tissue to identify genes that are differentially expressed between healthy individuals and patients with periodontitis.

## Research Question
Which genes are differentially expressed in gingival tissue from periodontitis patients compared with healthy controls?

## Data Acquisition
The dataset used is GSE173078 from GEO. It includes RNA-seq profiles from healthy, gingivitis, and periodontitis gingival tissue samples.

## Methods
Raw count data will be analyzed using DESeq2. Healthy and periodontitis samples will be compared to identify differentially expressed genes.

```{r}
res <- results(
  dds,
  contrast=c("condition",
             "Periodontitis",
             "Healthy")
)
```
```{r}
Filter 
sig <- subset(
  res,
  padj < 0.05 &
  abs(log2FoldChange) > 1
)
```



## Results
The main outputs will include:
- PCA plot
- MA plot
- Volcano plot
- Table of top differentially expressed genes

## Interpretation
Genes with increased expression in periodontitis may reflect inflammation, immune-cell recruitment, and periodontal tissue breakdown.

## Limitations
This analysis uses publicly available data and does not include clinical covariates such as age, sex, smoking status, or disease severity.

## Reproducibility
Session information and all code will be included.
