---
title: "HW 3 BMDA"
author: "Nat Lim"
date: "2026-06-13"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

In Homework 1 you downloaded and QC'd the 33 FASTQ files for the HIV/TB dataset from SRA. In this assignment you will carry those reads through a complete RNA-seq differential expression analysis.

## RNA-sequencing analysis

##### 1. Align the HIV/TB reads (downloaded in Homework 1) to the human genome reference using your choice of the `Rsubread` or the `STAR` aligners. Confirm whether or not you were able to successfully complete this step (on your honor).
   
  1. Yes
  2. No

```{r, eval=FALSE}
require("BiocManager")

library(Rsubread)
buildindex("genomefile", "genomefile.fa")

align("genomefile", "myfastqfile.fq", 
      output_file="myfastq.bam", nthreads = 4)
```


##### 2. Use the `featureCounts` function in the `Rsubread` package to generate a counts file for this dataset. Upload your feature counts .txt file here.



For the rest of the assignment, please use the feature counts provided by Dr. Johnson in the lecture materials: {https://github.com/wevanjohnson/2026_Spring_BMDA/raw/refs/heads/main/lectures/week6/example_data/features_combined.txt}{features_combined.txt (link)}



##### 3. Generate a `SummarizedExperiment` object for your counts. The `colData` for these data are provided in the `homework2\_metadata.txt` file. What function is used to create a SummarizedExperiment object?
  
  1. **SummarizedExperiment(assays = list(counts = as.matrix(counts)), colData = metadata)**
  2. makeSummarizedExp(as.matrix(counts), metadata)
  3. seObject(counts, metadata)
  4. summarizeData(data = counts, metadata = metadata)
  5. buildExperiment(assay = as.matrix(counts), metadata = metadata)

```{r}
library(SummarizedExperiment)

counts <- read.delim("features_combined.txt", row.names = 1, check.names = FALSE)

metadata <- read.delim("hivtb_SRR_metadata.txt")

rownames(metadata) <- metadata$SampleID
metadata <- metadata[colnames(counts), ]

se <- SummarizedExperiment(assays = list(counts = as.matrix(counts)), colData = metadata
)

se
```


##### 4. Preprocess these data by removing `TB-HIV-ART` samples (should be two of them), removing any genes with zero expression for all samples, and by generating a log counts per million assay. How may genes were removed with the zero expression filter?
 
1. 48
2. **2,204**
3. 9,700
4. 17,385

```{r}
se <- se[, se$disease_status != "TB-HIV-ART"]
dim(se)

keep <- rowSums(assay(se, "counts")) > 0
n_removed <- sum(!keep)
n_removed

se <- se[keep, ]
dim(se)

library(edgeR)
logcpm <- cpm(assay(se, "counts"), log = TRUE)
assay(se, "logCPM") <- logcpm

se
```


##### 5. Create a batch corrected assay in your `SummarizedExperiment` using ComBat-Seq. You can use the `ComBat\_Seq` function in the `sva` package, or simply do it in `BatchQC` and then extract the `SummarizedExperiment`. For this example, pretend that `disease\_status` is the batch variable. Note that this will remove the disease status variability, so **don't** use this assay in the following analyses! This was merely a practice for cases where you have an actual batch variable. What command/syntax would work to apply ComBat-Seq?
  
  1. **ComBat_seq(as.matrix(assay(se, "counts")), batch = se$disease_status)**
  2. ComBat_seq(counts = assay(se), batch = disease_status)
  3. BatchFix(se, "counts", batch = "disease_status")
  4. combatCorrect(se$counts, batch = se\$disease_status)
  5. ComBat(se, "counts", batch = "disease_status")
  
```{r, eval=FALSE}
library(sva)

combat_counts <- ComBat_seq(as.matrix(assay(se, "counts")), batch = se$disease_status)

assay(se, "combat_counts") <- combat_counts

se
```



##### 6. Apply PCA and UMAP to your data and generate dimension reduction plots for the results. Color the TB-HIV to the HIV only samples in different colors. Note that you should be using the log CPM values for this analysis. Please attach your UMAP plot here. 

```{r}
library(umap)

logcpm_mat <- t(assay(se, "logCPM"))

pca <- prcomp(logcpm_mat, scale. = TRUE)
pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  disease_status = se$disease_status
)

library(ggplot2)
ggplot(pca_df, aes(PC1, PC2, color = disease_status)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "PCA of log CPM values")

set.seed(42)
umap_res <- umap(logcpm_mat)
umap_df <- data.frame(
  UMAP1 = umap_res$layout[,1],
  UMAP2 = umap_res$layout[,2],
  disease_status = se$disease_status
)

ggplot(umap_df, aes(UMAP1, UMAP2, color = disease_status)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "UMAP of log CPM values")
```


##### 7. Use `DESeq2` to do a differential expression analysis (on the counts) comparing the TB-HIV to the HIV only samples. What is the twenty second gene on the differentially expressed gene list?
  
  1. IGF2BP3
  2. AIM2
  3. IL1R2
  4. RAB20
  5. **FCGR1C**

```{r}
library(DESeq2)

dds <- DESeqDataSetFromMatrix(
  countData = assay(se, "counts"),
  colData   = colData(se),
  design    = ~ disease_status
)

dds <- DESeq(dds)
res <- results(dds, contrast = c("disease_status", "TB-HIV", "HIV"))

# Order by significance (typically by p-value)
res_ordered <- res[order(res$pvalue), ]
head(as.data.frame(res_ordered), 25)

# The 22nd gene:
rownames(res_ordered)[22]
```



##### 8. Now conduct the same analysis using `limma` on the log CPM values. How do the `DESeq2` results compare to the `limma` results? What is the seventh gene on the differentially expressed gene list?
  
  1. IGF2BP3
  2. AIM2
  3. **IL1R2**
  4. RAB20
  5. FCGR1C

```{r}
library(limma)

logcpm_mat <- assay(se, "logCPM")  # genes as rows, samples as columns

design <- model.matrix(~ disease_status, data = colData(se))

fit <- lmFit(logcpm_mat, design)
fit <- eBayes(fit)

res_limma <- topTable(fit, coef = 2, number = Inf, sort.by = "P")
head(res_limma, 25)

# The 7th gene:
rownames(res_limma)[7]
```

##### 9. Give a heatmap plot of either the `DESeq2` or the `limma` results (top 50). Add a colorbar for disease status. Upload your plot here.

```{r}
library(ComplexHeatmap)
library(circlize)

# Top 50 genes from DESeq2 results
top50 <- rownames(res_ordered)[1:50]

# Pull their logCPM values
mat <- assay(se, "logCPM")[top50, ]

# Scale by row (z-score) so the heatmap shows relative expression patterns
mat_scaled <- t(scale(t(mat)))

# Annotation colorbar for disease status
ha <- HeatmapAnnotation(
  disease_status = se$disease_status,
  col = list(disease_status = c("TB-HIV" = "firebrick", "HIV" = "steelblue"))
)

Heatmap(
  mat_scaled,
  name = "z-score",
  top_annotation = ha,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_title = "Top 50 DESeq2 DE Genes (TB-HIV vs HIV)",
  row_names_gp = grid::gpar(fontsize = 6)
)
```


##### 10. Conduct a pathway analysis of the top 50 Limma genes using enrichR (through R or online). Use the "Reactome\_Pathways\_2024" database. What is the top scoring pathway?
   
  1. "Interferon Gamma Signaling"
  2. "Fatty Acid Metabolism"
  3. **"Cytokine Signaling in Immune System"**
  4. "RAB Geranylgeranylation"

```{r}
require("enrichR")  # if not already installed
library(enrichR)

# Confirm the database is available
dbs <- listEnrichrDbs()
"Reactome_Pathways_2024" %in% dbs$libraryName

# Top 50 limma genes
top50_limma <- rownames(res_limma)[1:50]

# Run enrichment
enriched <- enrichr(top50_limma, databases = "Reactome_Pathways_2024")

# View top results
reactome_results <- enriched[["Reactome_Pathways_2024"]]
head(reactome_results[order(reactome_results$P.value), c("Term","P.value","Adjusted.P.value")], 50)

# Top scoring pathway:
reactome_results$Term[which.min(reactome_results$P.value)]
```


##### 11. Conduct a TBSignatureProfiler analysis on these data -- including signature heatmaps, individual boxplots, and AUC boxplots. Interpret your findings. Upload one of your TBSignatureProfiler plots here.

```{r}
library(TBSignatureProfiler)

# Run profiling using ssGSEA on the logCPM assay
sigs_result <- runTBsigProfiler(
  input        = se,
  useAssay     = "logCPM",
  algorithm    = "ssGSEA",
  signatures   = TBsignatures,   # the full built-in signature library
  parallel.sz  = 1,
  combineSigAndAlgorithm = TRUE
)

# Check what signature score columns got added
colnames(colData(sigs_result))
```
```{r}
sigs_to_plot <- c("Chen_HIV_4", "Sambarey_HIV_10", "Sweeney_OD_3", "Zak_RISK_16")

signatureHeatmap(
  sigs_result,
  signatureColNames = sigs_to_plot,
  annotationColNames = "disease_status",
  scale = TRUE,
  showColumnNames = FALSE,
  name = "TB/HIV Signature Scores"
)
```

```{r}
signatureBoxplot(
  inputData = sigs_result,
  name = "Chen_HIV_4 Signature Score",
  signatureColNames = "Chen_HIV_4",
  annotationColName = "disease_status"
)
```

```{r}
auc_table <- tableAUC(
  SE_scored = sigs_result,
  annotationColName = "disease_status",
  signatureColNames = sigs_to_plot,
  num.boot = 100
)

auc_table
```

